Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2022-03-15 ms.tex
Syncing checkpoint before starting work on reviewer comments

We now estimate the scaling of $H_{\rm bc}$ with $\tau_0$. In the limit $J_{\rm bc} \gg H_{\rm bc}$, we find that $J_{\rm bc} \approx \sqrt{3} H_0$ from Equation (\ref{eq:bc2}). We  estimate $H_{\rm bc}$ from $J_{\rm bc}$ using Equation (\ref{eq:Hbc}) as
\be
H_{\rm bc}(R, \sigma) \approx \frac{1}{\sqrt{3}k\phi}\frac{ks}{\Delta}H_0\ {\sim}\ H_0 s \frac{x^2}{a}\ {\sim}\ H_0 \frac{1}{\tau_0}\frac{(a\tau_0)^{2/3}}{a}\ {\sim}\ H_0 (a\tau_0)^{-1/3},
\ee
where we have used $s\ {\sim}\ 1/\sigma\ {\sim}\ 1/\tau_0$ so that
\be \label{eq:hbc_scaling}
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto (a\tau_0)^{-1/3}.
\ee
At large $\tau_0$, it is expected that the correction term will be small, but it will become increasingly important as $\tau_0$ decreases. Our solution of the transfer equation is only valid when the peaks of the spectral energy distribution lie well outside of the Doppler core, e.g., for large $\tau_0$. The value of $x$ at which the Doppler and Lorentzian components of the line profile are equal is $x_{\rm cw}=3.3$. Setting $x_{\rm cw} = x_{\rm peak}$ and solving for $\tau_0$ gives the value at which the peak of the spectrum falls at the Doppler core boundary, which is $\tau_{\rm cp} \approx 10^5$ (``core-peak'' optical depth). Hence $H_{\rm bc}/H_0 {\sim} (\tau_{\rm cp}/\tau_0)^{1/3}$ is large at $\tau_0 \leq \tau_{\rm cp}$ and decreases relatively slowly as $\tau_0$ increases. Additionally, the optical depth at $x_{\rm peak}$ is proportional to $(a\tau_0)^{1/3}$, so photons here become optically thin when $a\tau_0 {\sim} 1$.

@@

 \begin{figure}
    \centering
    \includegraphics[width=\textwidth]{xinit_threepanel.pdf}
    \caption{The same as Figure \ref{fig:sol_mc_residual_0}, but for $\rm x_s = 0.0$ (top panel), $6.0$ (middle panel), and $12.0$ (bottom panel). The optical depth at each of these source frequencies is $\tau_s = 10^7, 77,$ and $19$, respectively.} 
    \label{fig:sol_mc_xinit}
\end{figure}

@@

In Figure \ref{fig:steadystate}, the fluence for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated''), and is compared with two analytic solutions: the steady-state $H_0$ solution (Equation \ref{eq:H0surf}) with no time dependence (labelled ``Steady State''), and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in the first line of Equation (\ref{eq:H0surf}). Additional spatial modes $n$ increase the solutions' accuracy in the core of the line. If more spatial modes are included, the agreement with the steady-state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x=x_s=0$. If additional frequency modes are included, faster-oscillating terms are incorporated into the Fourier sum over eigenmodes which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would primarily affect the distribution at late times due to the large mean diffusion time of photons near the line core. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated.

@@

Our characterization of the escape time distribution leads to a possible application of this work. Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a method to accelerate the radiative transfer calculation.

@@

%have found that even in this case of emission away from the line center, the eigenfunctions are symmetric about the line center. This is because Equation (\ref{eq:wkb_J_far}) has dropped a term, $Bi(x)$, that is infinite far from line center; yet, in the region $0 < \sigma < \sigma_s$, this term is finite and contributes an asymmetric piece to the solution. This term is not included in the Equations (\ref{eq:wkb_J}) and (\ref{eq:wkb_J_far}), meaning any solution produced by a sum over eigenmodes will always be symmetric about the line-center since the eigenfunctions themselves are symmetric even when $\sigma_s \neq 0$.

>>>>

We now estimate the scaling of $H_{\rm bc}$ with $\tau_0$. In the limit $J_{\rm bc} \gg H_{\rm bc}$, we find that $J_{\rm bc} \approx \sqrt{3} H_0$ from Equation (\ref{eq:bc2}). We  estimate $H_{\rm bc}$ from $J_{\rm bc}$ using Equation (\ref{eq:Hbc}) as
\be
H_{\rm bc}(R, \sigma) \approx \frac{1}{\sqrt{3}k\phi}\frac{ks}{\Delta}H_0\ {\sim}\ H_0 s \frac{x^2}{a}\ {\sim}\ H_0 \frac{1}{\tau_0}\frac{(a\tau_0)^{2/3}}{a}\ {\sim}\ H_0 (a\tau_0)^{-1/3},
\ee
where we have used $s\ {\sim}\ 1/\sigma\ {\sim}\ 1/\tau_0$ so that
\be \label{eq:hbc_scaling}
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto (a\tau_0)^{-1/3}.
\ee
At large $\tau_0$, it is expected that the correction term will be small, but it will become increasingly important as $\tau_0$ decreases. Our solution of the transfer equation is only valid when the peaks of the spectral energy distribution lie well outside of the Doppler core, i.e., for large $\tau_0$. The value of $x$ at which the Doppler and Lorentzian components of the line profile are equal is $x_{\rm cw}=3.3$. Setting $x_{\rm cw} = x_{\rm peak}$ and solving for $\tau_0$ gives the value at which the peak of the spectrum falls at the Doppler core boundary, which is $\tau_{\rm cp} \approx 10^5$ (``core-peak'' optical depth). Hence $H_{\rm bc}/H_0 {\sim} (\tau_{\rm cp}/\tau_0)^{1/3}$ is large at $\tau_0 \leq \tau_{\rm cp}$ and decreases relatively slowly as $\tau_0$ increases. Additionally, the optical depth at $x_{\rm peak}$ is proportional to $(a\tau_0)^{1/3}$, so photons here become optically thin when $a\tau_0 {\sim} 1$.

@@

 \begin{figure}
    \centering
    \includegraphics[width=\textwidth]{xinit_threepanel.pdf}
    \caption{The same as Figure \ref{fig:sol_mc_residual_0}, but for $\tau_0=10^7$ and $\rm x_s = 0.0$ (top panel), $6.0$ (middle panel), and $12.0$ (bottom panel). The optical depth at each of these source frequencies is $\tau_s = 10^7, 77,$ and $19$, respectively.} 
    \label{fig:sol_mc_xinit}
\end{figure}

@@

In Figure \ref{fig:steadystate}, the fluence for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated''), and is compared with two analytic solutions: the steady-state $H_0$ solution (Equation \ref{eq:H0surf}) with no time dependence (labelled ``Steady State''), and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in the first line of Equation (\ref{eq:H0surf}). Additional spatial modes $n$ increase the solutions' accuracy in the core of the line. If more spatial modes are included, the agreement with the steady-state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x=x_s=0$. If additional frequency modes are included, faster-oscillating terms are incorporated into the Fourier sum over eigenmodes which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would primarily affect the distribution at early times since the late time distribution is determined by the lowest order modes. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated.

@@

Our characterization of the escape time distribution leads to a possible application of this work. Models of the interaction of stellar \lya with the upper atmosphere of exoplanets and the associated transmission spectrum can be constructed with a treatment of resonant scattering in spherical geometry \citep{2017ApJ...851..150H}. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a method to accelerate the radiative transfer calculation.

@@
